Accuracy in Parameter Estimation and Simulation Approaches for Sample Size Planning with Multiple Stimuli

### 2023-09-08 manuscript_draft.Rmd
all the updates

---
title       : "Accuracy in Parameter Estimation and Simulation Approaches for Sample Size Planning with Multiple Stimuli"
shorttitle    : "AIPE SIMULATION"
author: 
 - name     : "Erin M. Buchanan"
  affiliation  : "1"
  corresponding : yes  # Define only one corresponding author
  address    : "326 Market St., Harrisburg, PA, 17101"
  email     : "ebuchanan@harrisburgu.edu"
  role:     # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
   - Conceptualization
   - Writing - Original Draft Preparation
   - Writing - Review & Editing
   - Resources
   - Validation 
   - Visualization 
   - Project Administration 
   - Formal Analysis
 - name     : "Other Folks as Per Order on Doc"
  affiliation  : "2"
  role:
   - Writing - Review & Editing
   - Data Curation
   - Resources
affiliation:
 - id      : "1"
  institution  : "Harrisburg University of Science and Technology"
 - id      : "2"
  institution  : "Other Instituions"
authornote: |
 The authors would like to thank K.D. Valentine for her assistance in formulation of correction scores. 
abstract: |
 The planning of the sample size for research studies often focuses on obtaining a significant result given a specified level of power, significance, and an anticipated effect size. This planning requires prior knowledge of the study design and a statistical analysis to calculate the proposed sample size. However, there may not be one specific testable analysis from which to derive power [@silberzahn2018many] or a hypothesis to test for the project (e.g., creation of a stimuli database). Modern power and sample size planning suggestions include accuracy in parameter estimation [AIPE, @kelley2007; @maxwell2008] and simulation of proposed analyses [@chalmers2020]. These toolkits provide flexibility in traditional power analyses that focus on the if-this, then-that approach, yet, both AIPE and simulation require either a specific parameter (e.g., mean, effect size, etc.) or statistical test for planning sample size. In this tutorial, we explore how AIPE and simulation approaches can be combined to accommodate studies that may not have a specific hypothesis test or wish to account for the potential of a multiverse of analyses. Specifically, the examples focus on studies that adopt multiple items and suggest that sample sizes can be planned to measure those items adequately and precisely, regardless of statistical test. We demonstrate that pilot data can be used to determine a sample size that represents well-measured data. This tutorial also provides multiple code vignettes that researchers can adapt and apply to their own measures.

keywords     : "accuracy in parameter estimation, power, sampling, simulation, hypothesis testing"
floatsintext   : no
figurelist    : no
tablelist     : no
footnotelist   : no
linenumbers    : yes
mask       : no
draft       : no
documentclass   : "apa6"
classoption    : "man"
output      : papaja::apa6_docx
bibliography: references.bib
---

```{r setup, include = FALSE}
set.seed(895893)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
library(dplyr)
library(ggplot2)
library(tidyr)
library(broom)
library(papaja)
library(rio)
library(nlme)
library(MuMIn)
library(pwr)
library(latex2exp)
```

@@

A recent review of power analyses found - across psychology journal articles - researchers did not provide enough information to understand their power analyses and often chose effect sizes that were unjustified [@beribisky2019]. One solution to this power analysis problem is the plethora of tools made available for researchers to make power computations accessible to non-statisticians with the caveat that a solid education in power is necessary to use these tools. G\*Power is one of the most popular free power software options [@faul2007; @erdfelder1996] that provides a simple point and click graphical user interface for power calculations [however, see @brysbaert]. Web-based tools have also sprung up for overall and statistical test specific sample size planning including <https://powerandsamplesize.com> and <https://designingexperiments.com> [@anderson2017]. *R*-coding based packages, such as *pwr* [@champely2017], *faux* [@debruine2021], and *SimDesign* [@chalmers2020], can be used to examine power and sample size planning usually with simulation. Researchers must be careful using any toolkit, as errors can occur with the over-reliance on software [e.g., it should not be a substitute for critical thinking, @nuijten2016]. When computing sample size estimates, it is important to remember that these values are estimations, not exact calculations guaranteed to produce a specific result [@batterham2005]. For example, if any parameter estimated by the researcher was not found in the study (i.e., a smaller effect size than used for the power analysis), then the sample size estimate tied to that specific level of power may be incorrect.

@@

2)  Calculate the standard error (SE) in the subset of pilot data associated with each item to create a cut-off score that defines when items are considered "accurately measured". Next, the simulations below will explore what criterion should be used to determine the cut-off score from the pilot data.

@@

5)  Find the minimum sample size so that 80%, 85%, 90%, and 95% of the items meet the cut-off score and can be considered accurately measured. We recommend these scores to ensure that most items are accurately measured, in a similar vein to the common power-criterion suggestions. Each researcher can determine which of these is their minimum or maximum sample size (e.g., individuals can choose to use 80% as a minimum and 90% as a maximum or use values from Step 3 based on resources).

6)  Report these values, and designate a minimum sample size, the cut-off criterion, and the maximum sample size. Each researcher should also report if they plan to use an adaptive design, which would stop data collection after meeting the cut-off criterion for each item.

These steps are summarized in Table \@ref(tab:table-summary) 1. We will first demonstrate the ideas behind the steps using open data [@brysbaert2014; @balota2007]. This example will reveal a few areas of needed exploration for the steps. Next, we portray simulations for the proposed procedure and find solutions to streamline and improve the sample size estimation procedure. Finally, we include additional resources for researchers to use to implement the estimation procedure.

```{r table-summary, results = 'asis'}
apa_table(data.frame(
  "Step Number" = c("1", "2" , "3", "4", "5", "6"),
  "Proposed Steps" = c(
    "Use representative pilot data.",
    "Calculate standard error of each of the items in the pilot data. Determine the appropriate SE for the stopping rule.",
    "Create bootstrapped samples of your pilot data starting with at least 20 participants up to a maximum number of participants.", 
    "Calculate the standard error of each of the items in the bootstrapped data. From these scores, calculate the percent of items below the cut-off score from Step 2.",
    "Determine the sample size at which 80%, 85%, 90%, 95% of items are below the cut-off score.",
    "Report all values. Designate one as the minimum sample size, the cut-off score as the stopping rule for adaptive designs, and the maximum sample size.",
    "Updated Steps" = c("Use representative pilot data.",
    "Calculate standard error of each of the items in the pilot data. Using the 40%, determine the cut-off and stopping rule for the standard error of the items.",
    "Create bootstrapped samples of your pilot data starting with at least 20 participants up to a maximum number of participants.", 
    "Calculate the standard error of each of the items in the bootstrapped data. From these scores, calculate the percent of items below the cut-off score from Step 2.",
    "Determine the sample size at which 80%, 85%, 90%, 95% of items are below the cut-off score. Use the correction formula to adjust your proposed sample size based on pilot data size, power, and percent variability.",
    "Report all values. Designate one as the minimum sample size, the cut-off score as the stopping rule for adaptive designs, and the maximum sample size.")
  ), 
  caption = "Proposed Procedure for Powering Studies with Multiple Items"
))
```

@@

The second dataset includes a large scale dataset with response latencies, the English Lexicon Project [@balota2007]. The English Lexicon Project consists of lexical decision response latencies for English words. In a lexical decision task, participants simply select "word" for real words (e.g., *dog*) and "nonword" for pseudowords (e.g., *wug*). The trial level data is available here [<https://elexicon.wustl.edu/>]. Critically, in each of these examples, the individual trial level data for each item is available to simulate and calculate standard errors on. Data that has been summarized could potentially be used, as long as the original standard deviations for each item were present. From the mean and standard deviation for each item, a simulated pilot dataset could be generated for estimating new sample sizes. All code to estimate sample sizes is provided on our OSF page.

@@

*Step 1*. The concreteness ratings data includes `r length(unique(conDF$Word))` concepts that were rated for their concreteness. We randomly selected *n* = 20 abstract words ($M_{Rating}$ \<= 2) and *n* = 20 concrete words ($M_{Rating}$ \>= 4). In the original study, not every participant rated every word, which created uneven sample sizes for each word. Further, participants were allowed to indicate they did not know a word, and this data was set to missing data. In our sample of 40 words, the average pilot sample size was `r round(mean(tapply(con_small$Rating, con_small$Word, length)), digits = 2)` (*SD* = `r round(sd(tapply(con_small$Rating, con_small$Word, length)), digits = 2)`), and we will use `r pilot_size` as our pilot sample size for this example.

@@

*Step 2*. Table \@ref(tab:table-example) demonstrates the cut-off scores for deciles of the standard errors for the items for the concreteness ratings and lexical decision response latencies. A researcher could potentially pick any of these cut-offs or other percentage options not shown here. We will use simulation to determine the suggestion that best captures the balance of adequately powering our sample and feasibility.

*Step 3-5*. The pilot data was then simulated, with replacement, with samples from 20 to 300 increasing in units of 5, separately for concreteness and lexical decision times (Step 3). The standard error of each item was calculated for the bootstrapped samples (Step 4), and the percentage of items below each potential cut-off was gathered (Step 5). The smallest sample size with at least 80%, 85%, 90%, and 95% of items below the cut-off are reported in Table \@ref(tab:table-example) (Step 5).

@@

```{r table-example, results = 'asis'}
# concreteness before - elp before 
  # sample size
  # decile cut-offs? 
  # number of participants for each decile? 
  # 80, 85, 90, 95

@@

results_table_example <- as.data.frame(results_table)
colnames(results_table_example) <- c("Deciles", "C Decile SE", "C 80", "C 85", "C 90", "C 95",
             "L Decile SE", "L 80", "L 85", "L 90", "L 95")

@@

*Step 6*. In the last step, the researcher would indicate their smallest sample size, the cut-off standard error criterion if they wanted to adaptively test (e.g., examine the SE after each participant and stop data collection if all items reached criteria), and their maximum sample size. As mentioned earlier, the decile for a balanced standard error cut-off is unclear and without guidance, a potential set of researcher degrees of freedom [CITE]. Even though both scales appear to converge on similar sample size suggestions for each decile and percent level, the impact of scale size (i.e., 1-5 versus 0-`r max(elp_small$RT, na.rm = T)`) and heterogeneity of item standard errors (concrete $SD_{SD}$ = `r apa_num(sd_items_c)` and lexical $SD_{SD}$ = `r apa_num(sd_items_e)`) is not obvious. Last, by selecting the ends of the distribution for our concreteness words, skew of the distribution may additionally impact our estimates. Each of these will be explored in our simulation.

@@

In order to evaluate our approach, we used data simulation to create representative pilot datasets of several popular cognitive scales (1-7 measurements, 0-100 percentage measurements, and 0-3000 response latency scale data). For each of these scales, we also manipulated item heterogeneity by simulating small differences in item variances to large differences in item variances based on original scale size. On each of the simulated datasets, we applied the above proposed method to determine how the procedure would perform and evaluate what criteria should be used for cut-off selection (Step 2). This procedure was performed on distributions in the middle of the scale (i.e., normal) and at the ceiling of the scale (i.e., skewed). With this simulation, we will answer several questions:

1)  How do pilot data influence sample size suggestions? A. How does scale size impact sample size estimations? In theory, the size of the scale used should not impact the power estimates; however, larger scales have a potential for more variability in their item standard deviations (see point C). B. How does distribution skew impact sample size estimations? Skew can potentially decrease heterogeneity (i.e., all items are at ceiling, and therefore, variance between item standard errors is low) or could increase heterogeneity (i.e., some items are skewed, while others are not). Therefore, we expect skew to impact the estimates in the same way as point C. C. How does heterogeneity impact sample size estimations? Heterogeneity should decrease power [@alexander1994; @rheinheimer2001], and thus, increased projected sample sizes should be proposed as heterogeneity of item variances increases.
2)  Do the results match what one might expect for traditional power curves? Power curves are asymptotic, that is, they "level off" as sample size increases. Therefore, we expect that our procedure should also demonstrate a leveling off effect as pilot data sample size increases. For example, if one has a 500-person pilot study, our simulations should suggest a point at which items are likely measured well, which may have happened well before 500.
3)  What should the suggested cut-off standard error decile be?

@@

```{r table-sim, results = 'asis'}
apa_table(data.frame(
  Information = c("Minimum", "Maximum", "$\\mu$", "$Skewed \\mu$", "$\\sigma_{\\mu}$", 
                  "$\\sigma$", "Small $\\sigma_{\\sigma}$", 
                  "Medium $\\sigma_{\\sigma}$", "Large $\\sigma_{\\sigma}$"),
  Likert = c(1, 7, 4, 6, .25, 2, .2, .4, .8),
  Percent = c(0, 100, 50, 85, 10, 25, 4, 8, 16),
  Milliseconds = c(0, 3000, 1000, 2500, 150, 400, 50, 100, 200)
), caption = "Parameter Values for Data Simulation"
)

@@

In this section, we simulate what a researcher might do if they follow our suggested application of AIPE to sample size planning based on well measured items. Assuming that each pilot sample represents a dataset that a researcher has collected (Step 1), the standard errors for each item were calculated to mimic the AIPE procedure of finding an appropriately small confidence interval, as SE functions as the main component of the formula for normal distribution confidence intervals. Standard errors were calculated at each decile of the items up to 90% (i.e., 0% smallest SE, 10% ..., 90% largest SE). The lower deciles would represent a strict criterion for accurate measurement, as many items would need smaller SEs to meet cut-off scores, while the higher deciles would represent less strict criteria for cut-off scores.

@@

Next, the percentage of items that fall below the cut-off scores and therefore would be considered "well-measured" were calculated for each decile by sample (Step 4). From these data, we pinpoint the smallest suggested sample size at which 80%, 85%, 90% and 95% of the items fall below the cut-off criterion (Step 5). These values were chosen as popular measures of "power" in which one could determine the minimum suggested sample size (potentially 80% of the items) and the maximum suggested sample size (selected from a higher percentage, such as 90% or 95%).

@@

Figure \@ref(fig:scale-size-figure) demonstrates the influence of scale size on the results separate by potential cut-off level. The black dots denote the original sample size for reference. Larger scales have more potential variability, and therefore, we see that percent and millisecond scales project a larger required sample size. This relationship does not appear to be linear with scale size, as percent scales often represent the highest projected sample size. Potentially, this finding is due to the larger proportion of possible variance -- the variance of the item standard deviations / total possible variance -- was largest for percent scales in this set of simulations ($p_{Percent}$ = `r apa_num(mean_var["Percent"], gt1 = FALSE)`. This finding may be an interaction with heterogeneity, as the Likert scale had the next highest percent variability in item standard errors ($p_{Likert}$ = `r apa_num(mean_var["Likert"], gt1 = FALSE)`, followed by milliseconds ($p_{Milliseconds}$ = `r apa_num(mean_var["Milliseconds"], gt1 = FALSE)`.

@@

```{r scale-hetero-figure, echo = F, fig.cap="Add a good caption here."}
ggplot(summary_long %>% 
         group_by(original_n, power, variability) %>% 
         summarize(sample_size_avg = mean(sample_size)),
       aes(original_n, sample_size_avg, color = variability)) + 
  geom_point() + 
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  theme_linedraw() + 
  xlab("Pilot Sample Size") + 
  ylab("Projected Sample Size") + 
  scale_color_discrete(name = "Item Heterogeneity") +
  facet_wrap(~power)
```

```{r strongest-predict}
regression_DF <- summary_long %>% 
  group_by(original_n, scale_size, variability, data_type) %>% 
  summarize(sample_size = mean(sample_size),
            sd_item = mean(sd_item),
            percent_var = mean(percent_var))
model.samplesize <- lm(sample_size ~ original_n + scale_size + percent_var + data_type,
                       data = regression_DF)
summary(model.samplesize, correlation = TRUE)
```

@@

apa_table(
  temp_table, 
  col.names = c("Term", "Estimate", "$SE$", "$t$", "$p$", "$pr^2$"),
  escape = FALSE
)
```

@@

In our second question, we examined if the suggested procedure was sensitive to the amount of information present in the pilot data. Larger pilot data is more informative, and therefore, we should expect a lower projected sample size. As shown in each figure presented already, we do not find this effect. These simulations from the pilot data would nearly always suggest a larger sample size - mostly in a linear trend increasing with sample sizes. This result comes from the nature of the procedure - if we base our estimates on a SE cut-off, we will almost always need a bit more people for items to meet those goals. This result does not achieve our second goal.

@@

```{r corrected-figure, echo = F, fig.cap="A corrected figure update this caption."}
decay <- 1-sqrt((summary_long$original_n-20)/summary_long$original_n)^log2(summary_long$original_n)
summary_long$new_sample <- summary_long$sample_size*decay
ggplot(summary_long %>% 
         group_by(original_n, power, variability) %>% 
         summarize(sample_size_avg = mean(sample_size),
                   new_sample_size_avg = mean(new_sample)),
       aes(original_n, new_sample_size_avg, color = variability)) + 
  geom_point() + 
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  theme_linedraw() + 
  xlab("Pilot Sample Size") + 
  ylab("Projected Sample Size") + 
  scale_color_discrete(name = "Item Heterogeneity") +
  facet_wrap(~power)
```

@@

```{r table-scores-updated, results = 'asis'}
c_sample <- c(results_table_example$`C 80`[4], 
              results_table_example$`C 85`[4],
              results_table_example$`C 90`[4],
              results_table_example$`C 95`[4])

@@

apa_table(table_correction_applied, 
          caption = "Applied Correction for Each Proposed Sample Size")
```

>>>>

---
title       : "Accuracy in Parameter Estimation and Simulation Approaches for Sample Size Planning with Multiple Stimuli"
shorttitle    : "AIPE BOOTSTRAPPING"
author: 
- name     : "Erin M. Buchanan"
  affiliation  : "1"
  corresponding : yes  # Define only one corresponding author
  address    : "326 Market St., Harrisburg, PA, 17101"
  email     : "ebuchanan@harrisburgu.edu"
  role:     # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
   - Conceptualization
   - Writing - Original Draft Preparation
   - Writing - Review & Editing
   - Resources
   - Validation 
   - Visualization 
   - Project Administration 
   - Formal Analysis
- name     : "Other Folks as Per Order on Doc"
  affiliation  : "2"
  role:
   - Writing - Review & Editing
   - Data Curation
   - Resources
affiliation:
- id      : "1"
  institution  : "Harrisburg University of Science and Technology"
- id      : "2"
  institution  : "Other Instituions"
authornote: |

abstract: |
 The planning of the sample size for research studies often focuses on obtaining a significant result given a specified level of power, significance, and an anticipated effect size. This planning requires prior knowledge of the study design and a statistical analysis to calculate the proposed sample size. However, there may not be one specific testable analysis from which to derive power [@silberzahn2018many] or a hypothesis to test for the project (e.g., creation of a stimuli database). Modern power and sample size planning suggestions include accuracy in parameter estimation [AIPE, @kelley2007; @maxwell2008] and simulation of proposed analyses [@chalmers2020]. These toolkits provide flexibility in traditional power analyses that focus on the if-this, then-that approach, yet, both AIPE and simulation require either a specific parameter (e.g., mean, effect size, etc.) or statistical test for planning sample size. In this tutorial, we explore how AIPE and simulation approaches can be combined to accommodate studies that may not have a specific hypothesis test or wish to account for the potential of a multiverse of analyses. Specifically, the examples focus on studies that adopt multiple items and suggest that sample sizes can be planned to measure those items adequately and precisely, regardless of statistical test. We demonstrate that pilot data can be used to determine a sample size that represents well-measured data. This tutorial also provides multiple code vignettes that researchers can adapt and apply to their own measures.

keywords     : "accuracy in parameter estimation, power, sampling, simulation, hypothesis testing"
floatsintext   : no
figurelist    : no
tablelist     : no
footnotelist   : no
linenumbers    : yes
mask       : no
draft       : no
classoption    : "man"
bibliography: references.bib
header-includes:
  - |
    \makeatletter
    \renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
      {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
      {-1em}%
      {\normalfont\normalsize\bfseries\typesectitle}}

    \renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
      {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
      {-\z@\relax}%
      {\normalfont\normalsize\bfseries\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
    \makeatother

csl               : "`r system.file('rmd', 'apa7.csl', package = 'papaja')`"
documentclass     : "apa7"
output            : papaja::apa6_pdf
---

```{r setup, include = FALSE, echo = FALSE}
set.seed(895893)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
library(dplyr)
library(ggplot2)
library(tidyr)
library(broom)
library(papaja)
library(rio)
library(nlme)
library(MuMIn)
library(pwr)
library(latex2exp)
```

@@

A recent review of power analyses found - across psychology journal articles - researchers did not provide enough information to understand their power analyses and often chose effect sizes that were unjustified [@beribisky2019]. One solution to this power analysis problem is the plethora of tools made available for researchers to make power computations accessible to non-statisticians with the caveat that a solid education in power is necessary to use these tools. G\*Power is one of the most popular free power software options [@faul2007; @erdfelder1996] that provides a simple point and click graphical user interface for power calculations [however, see @brysbaert2019]. Web-based tools have also sprung up for overall and statistical test specific sample size planning including <https://powerandsamplesize.com> and <https://designingexperiments.com> [@anderson2017]. *R*-coding based packages, such as *pwr* [@champely2017], *faux* [@debruine2021], and *SimDesign* [@chalmers2020], can be used to examine power and sample size planning usually with simulation. Researchers must be careful using any toolkit, as errors can occur with the over-reliance on software [e.g., it should not be a substitute for critical thinking, @nuijten2016]. When computing sample size estimates, it is important to remember that these values are estimations, not exact calculations guaranteed to produce a specific result [@batterham2005]. For example, if any parameter estimated by the researcher was not found in the study (i.e., a smaller effect size than used for the power analysis), then the sample size estimate tied to that specific level of power may be incorrect.

@@

2)  Calculate the standard error (SE) in the subset of pilot data associated with each item to create a cutoff score that defines when items are considered "accurately measured". Next, the simulations below will explore what criterion should be used to determine the cutoff score from the pilot data.

@@

5)  Find the minimum sample size so that 80%, 85%, 90%, and 95% of the items meet the cutoff score and can be considered accurately measured. We recommend these scores to ensure that most items are accurately measured, in a similar vein to the common power-criterion suggestions. Each researcher can determine which of these is their minimum or maximum sample size (e.g., individuals can choose to use 80% as a minimum and 90% as a maximum or use values from Step 3 based on resources).

6)  Report these values, and designate a minimum sample size, the cutoff criterion, and the maximum sample size. Each researcher should also report if they plan to use an adaptive design, which would stop data collection after meeting the cutoff criterion for each item.

These steps are summarized in Table \@ref(tab:table-summary). We will first demonstrate the ideas behind the steps using open data [@brysbaert2014; @balota2007]. This example will reveal a few areas of needed exploration for the steps. Next, we portray simulations for the proposed procedure and find solutions to streamline and improve the sample size estimation procedure. Finally, we include additional resources for researchers to use to implement the estimation procedure.

```{r table-summary, results = 'asis'}
apa_table(data.frame(
  "Step Number" = c("1", "2" , "3", "4", "5", "6"),
  "Proposed Steps" = c(
    "Use representative pilot data.",
    "Calculate standard error of each of the items in the pilot data. Determine the appropriate SE for the stopping rule.",
    "Create bootstrapped samples of your pilot data starting with at least 20 participants up to a maximum number of participants.", 
    "Calculate the standard error of each of the items in the bootstrapped data. From these scores, calculate the percent of items below the cutoff score from Step 2.",
    "Determine the sample size at which 80%, 85%, 90%, 95% of items are below the cutoff score.",
    "Report all values. Designate one as the minimum sample size, the cutoff score as the stopping rule for adaptive designs, and the maximum sample size."),
    "Updated Steps" = c("Use representative pilot data.",
    "Calculate standard error of each of the items in the pilot data. Using the 40%, determine the cutoff and stopping rule for the standard error of the items.",
    "Create bootstrapped samples of your pilot data starting with at least 20 participants up to a maximum number of participants.", 
    "Calculate the standard error of each of the items in the bootstrapped data. From these scores, calculate the percent of items below the cutoff score from Step 2.",
    "Determine the sample size at which 80%, 85%, 90%, 95% of items are below the cutoff score. Use the correction formula to adjust your proposed sample size based on pilot data size, power, and percent variability.",
    "Report all values. Designate one as the minimum sample size, the cutoff score as the stopping rule for adaptive designs, and the maximum sample size.")
  ), 
  caption = "Proposed Procedure for Powering Studies with Multiple Items",
  col.names = c("Step", "Proposed Steps", "Updated Steps"),
  align = c("m{1cm}", rep("m{6cm}", 2)),
  font_size = "footnotesize"
)
```

@@

The second dataset includes a large scale dataset with response latencies, the English Lexicon Project [@balota2007]. The English Lexicon Project consists of lexical decision response latencies for English words. In a lexical decision task, participants simply select "word" for real words (e.g., *dog*) and "nonword" for pseudowords (e.g., *wug*). The trial level data is available here: <https://elexicon.wustl.edu/>. Critically, in each of these examples, the individual trial level data for each item is available to simulate and calculate standard errors on. Data that has been summarized could potentially be used, as long as the original standard deviations for each item were present. From the mean and standard deviation for each item, a simulated pilot dataset could be generated for estimating new sample sizes. All code to estimate sample sizes is provided on our OSF page, and this manuscript was created with a *papaja* [@aust2022] formatted Rmarkdown document.

@@

*Step 1*. The concreteness ratings data includes `r length(unique(conDF$Word))` concepts that were rated for their concreteness. We randomly selected *n* = 20 abstract words ($M_{Rating}$ \<= 2) and *n* = 20 concrete words ($M_{Rating}$ \>= 4). In the original study, not every participant rated every word, which created uneven sample sizes for each word. Further, participants were allowed to indicate they did not know a word, and this data was set to missing data. In our sample of 40 words, the average pilot sample size was `r round(mean(tapply(con_small$Rating, con_small$Word, length)), digits = 2)` (*SD* = `r round(sd(tapply(con_small$Rating, con_small$Word, length)), digits = 2)`), and we will use `r pilot_size_c` as our pilot sample size for this example.

@@

*Step 2*. Table \@ref(tab:table-example) demonstrates the cutoff scores for deciles of the standard errors for the items for the concreteness ratings and lexical decision response latencies. A researcher could potentially pick any of these cutoffs or other percentage options not shown here. We will use simulation to determine the suggestion that best captures the balance of adequately powering our sample and feasibility.

*Step 3-5*. The pilot data was then simulated, with replacement, with samples from 20 to 300 increasing in units of 5, separately for concreteness and lexical decision times (Step 3). The standard error of each item was calculated for the bootstrapped samples (Step 4), and the percentage of items below each potential cutoff was gathered (Step 5). The smallest sample size with at least 80%, 85%, 90%, and 95% of items below the cutoff are reported in Table \@ref(tab:table-example) (Step 5).

@@

```{r table-example, results = 'asis'}
# concreteness before - elp before 
  # sample size
  # decile cutoffs? 
  # number of participants for each decile? 
  # 80, 85, 90, 95

@@

results_table_example <- as.data.frame(results_table)
colnames(results_table_example) <- c("Deciles", "C Decile SE", "C 80", "C 85", "C 90", "C 95",
             "L Decile SE", "L 80", "L 85", "L 90", "L 95")
results_table_example$`C Decile SE` <- apa_num(as.numeric(results_table_example$`C Decile SE`), digits = 2)
results_table_example$`L Decile SE` <- apa_num(as.numeric(results_table_example$`L Decile SE`), digits = 2)

@@

*Step 6*. In the last step, the researcher would indicate their smallest sample size, the cutoff standard error criterion if they wanted to adaptively test (e.g., examine the SE after each participant and stop data collection if all items reached criteria), and their maximum sample size. As mentioned earlier, the decile for a balanced standard error cutoff is unclear and without guidance, a potential set of researcher degrees of freedom [CITE]. Even though both scales appear to converge on similar sample size suggestions for each decile and percent level, the impact of scale size (i.e., 1-5 versus 0-`r max(elp_small$RT, na.rm = T)`) and heterogeneity of item standard errors (concrete $SD_{SD}$ = `r apa_num(sd_items_c)` and lexical $SD_{SD}$ = `r apa_num(sd_items_e)`) is not obvious. Last, by selecting the ends of the distribution for our concreteness words, skew of the distribution may additionally impact our estimates. Each of these will be explored in our simulation.

@@

In order to evaluate our approach, we used data simulation to create representative pilot datasets of several popular cognitive scales (1-7 measurements, 0-100 percentage measurements, and 0-3000 response latency scale data). For each of these scales, we also manipulated item heterogeneity by simulating small differences in item variances to large differences in item variances based on original scale size. On each of the simulated datasets, we applied the above proposed method to determine how the procedure would perform and evaluate what criteria should be used for cutoff selection (Step 2). This procedure was performed on distributions in the middle of the scale (i.e., normal) and at the ceiling of the scale (i.e., skewed). With this simulation, we will answer several questions:

1)  How do pilot data influence sample size suggestions?

    A. How does scale size impact sample size estimations? In theory, the size of the scale used should not impact the power estimates; however, larger scales have a potential for more variability in their item standard deviations (see point C).

    B. How does distribution skew impact sample size estimations? Skew can potentially decrease heterogeneity (i.e., all items are at ceiling, and therefore, variance between item standard errors is low) or could increase heterogeneity (i.e., some items are skewed, while others are not). Therefore, we expect skew to impact the estimates in the same way as point C.

    C. How does heterogeneity impact sample size estimations? Heterogeneity should decrease power [@alexander1994; @rheinheimer2001], and thus, increased projected sample sizes should be proposed as heterogeneity of item variances increases.

2)  Do the results match what one might expect for traditional power curves? Power curves are asymptotic, that is, they "level off" as sample size increases. Therefore, we expect that our procedure should also demonstrate a leveling off effect as pilot data sample size increases. For example, if one has a 500-person pilot study, our simulations should suggest a point at which items are likely measured well, which may have happened well before 500.

3)  What should the suggested cutoff standard error decile be?

@@

```{r table-sim, results = 'asis'}
apa_table(data.frame(
  Information = c("Minimum", "Maximum", "$\\mu$", "$Skewed \\mu$", "$\\sigma_{\\mu}$", 
                  "$\\sigma$", "Small $\\sigma_{\\sigma}$", 
                  "Medium $\\sigma_{\\sigma}$", "Large $\\sigma_{\\sigma}$"),
  Likert = c(1, 7, 4, 6, .25, 2, .2, .4, .8),
  Percent = c(0, 100, 50, 85, 10, 25, 4, 8, 16),
  Milliseconds = c(0, 3000, 1000, 2500, 150, 400, 50, 100, 200)
), caption = "Parameter Values for Data Simulation",
escape = FALSE
)

@@

In this section, we simulate what a researcher might do if they follow our suggested application of AIPE to sample size planning based on well measured items. Assuming that each pilot sample represents a dataset that a researcher has collected (Step 1), the standard errors for each item were calculated to mimic the AIPE procedure of finding an appropriately small confidence interval, as SE functions as the main component of the formula for normal distribution confidence intervals. Standard errors were calculated at each decile of the items up to 90% (i.e., 0% smallest SE, 10% ..., 90% largest SE). The lower deciles would represent a strict criterion for accurate measurement, as many items would need smaller SEs to meet cutoff scores, while the higher deciles would represent less strict criteria for cutoff scores.

@@

Next, the percentage of items that fall below the cutoff scores and therefore would be considered "well-measured" were calculated for each decile by sample (Step 4). From these data, we pinpoint the smallest suggested sample size at which 80%, 85%, 90% and 95% of the items fall below the cutoff criterion (Step 5). These values were chosen as popular measures of "power" in which one could determine the minimum suggested sample size (potentially 80% of the items) and the maximum suggested sample size (selected from a higher percentage, such as 90% or 95%).

@@

Figure \@ref(fig:scale-size-figure) demonstrates the influence of scale size on the results separate by potential cutoff level. The black dots denote the original sample size for reference. Larger scales have more potential variability, and therefore, we see that percent and millisecond scales project a larger required sample size. This relationship does not appear to be linear with scale size, as percent scales often represent the highest projected sample size. Potentially, this finding is due to the larger proportion of possible variance -- the variance of the item standard deviations / total possible variance -- was largest for percent scales in this set of simulations ($p_{Percent}$ = `r apa_num(mean_var["Percent"], gt1 = FALSE)`). This finding may be an interaction with heterogeneity, as the Likert scale had the next highest percent variability in item standard errors ($p_{Likert}$ = `r apa_num(mean_var["Likert"], gt1 = FALSE)`), followed by milliseconds ($p_{Milliseconds}$ = `r apa_num(mean_var["Milliseconds"], gt1 = FALSE)`).

@@

```{r scale-hetero-figure, echo = F, fig.cap="Simulated pilot sample size compared to projected sample size for differing amounts of heterogeneity on each scale. "}
ggplot(summary_long %>% 
         group_by(original_n, power, variability) %>% 
         summarize(sample_size_avg = mean(sample_size)),
       aes(original_n, sample_size_avg, color = variability)) + 
  geom_point() + 
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  theme_linedraw() + 
  xlab("Pilot Sample Size") + 
  ylab("Projected Sample Size") + 
  scale_color_discrete(name = "Item Heterogeneity") +
  facet_wrap(~power)
```

```{r strongest-predict, include = F, echo = F}
regression_DF <- summary_long %>% 
  group_by(original_n, scale_size, variability, data_type) %>% 
  summarize(sample_size = mean(sample_size),
            sd_item = mean(sd_item),
            percent_var = mean(percent_var))
model.samplesize <- lm(sample_size ~ original_n + scale_size + percent_var + data_type,
                       data = regression_DF)
summary(model.samplesize, correlation = TRUE)
```

@@

apa_table(
  temp_table, 
  col.names = c("Term", "Estimate", "$SE$", "$t$", "$p$", "$pr^2$"),
  caption = "Prediction of Proposed Sample Size from Simulated Variables"
)
```

@@

In our second question, we examined if the suggested procedure was sensitive to the amount of information present in the pilot data. Larger pilot data is more informative, and therefore, we should expect a lower projected sample size. As shown in each figure presented already, we do not find this effect. These simulations from the pilot data would nearly always suggest a larger sample size - mostly in a linear trend increasing with sample sizes. This result comes from the nature of the procedure - if we base our estimates on a SE cutoff, we will almost always need a bit more people for items to meet those goals. This result does not achieve our second goal.

@@

```{r corrected-figure, echo = F, fig.cap="Corrected projected sample sizes for variability and power levels. "}
decay <- 1-sqrt((summary_long$original_n-20)/summary_long$original_n)^log2(summary_long$original_n)
summary_long$new_sample <- summary_long$sample_size*decay
ggplot(summary_long %>% 
         group_by(original_n, power, variability) %>% 
         summarize(sample_size_avg = mean(sample_size),
                   new_sample_size_avg = mean(new_sample)),
       aes(original_n, new_sample_size_avg, color = variability)) + 
  geom_point() + 
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  theme_linedraw() + 
  xlab("Pilot Sample Size") + 
  ylab("Projected Sample Size") + 
  scale_color_discrete(name = "Item Heterogeneity") +
  facet_wrap(~power)
```

@@

```{r table-scores-updated, results = 'asis', eval = F}
c_sample <- c(results_table_example$`C 80`[4], 
              results_table_example$`C 85`[4],
              results_table_example$`C 90`[4],
              results_table_example$`C 95`[4])

@@

apa_table(table_correction_applied, 
          caption = "Applied Correction for Each Proposed Sample Size", 
          font_size = "footnotesize")
```
